---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Group-wise normalization of microbiome data

<!-- badges: start -->
<!-- badges: end -->

## Overview
Differential abundance analysis (DAA) of the microbiome is statistically challenging
because 16S sequencing data are compositional. To account for this, many common 
DAA methods require calculating "normalization factors" that serve as the offset 
term (or similar) in the statistical model. DAA methods that require normalization include edgeR, 
DESeq2, and MetagenomeSeq. Some common normalization methods are TSS, CSS, TMM, 
RLE, GMPR, and Wrench. 

This repository provides companion code for our article, "Group-wise normalization
in differential abundance analysis of microbiome samples", which presented a 
novel normalization framework based on comparisons of group-level, rather
than sample-level, log fold changes. We performed a simulation study to compare 
the performance of the proposed framework with traditional normalization methods
that adopt a sample-wise approach.

Our simulation study encompassed two types of data: model-based data that was 
generated using a log-normal multinomial distribution, and synthetic data generated by
re-sampling microbiome data from two studies, PHACS and MLVS/MBS. These datasets 
cannot be shared publicly for confidentiality reasons, but we 
provide a script for generating fake versions of them in the `utils` 
folder. Our primary simulation study can be replicated using the scripts 
`model_based_simulations.R` and `synthetic_data_simulations.R` in the `analysis.R`
folder. The simulations vary factors such as the noise level, sample size, and 
percent of taxa that are differentially abundant, for a variety of normalization 
methods and DAA methods. The scripts 
`additional_sims.R` and `model_based_simulations_confounder.R` supplement
the primary simulations by comparing normalization methods in subsettings with
reduced compositional bias, imbalanced library sizes between groups, no
differentially abundant taxa, or adjustment for a confounding variable. The
results can be plotted using the script `process_output.R`.

The simulations performed in the article comprised 1,000 replications of each
setting, which is too much to be run locally or on a single computing core. 
Replicating the full study requires outsourcing simulations to a remote
computing cluster. To run a subset of the simulations locally, one could, for example:

1. Open the R Project file `microbiome_groupwise_normalization.Rproj` to start
RStudio in the proper working directory.
2. Ensure the packages listed in the script `utils/install_packages.R` are all installed.
2. Open the file `model_based_simulations.R`. 
3. Set the `total_reps` parameter to 100.
4. Set the `norms` (normalization methods), `methods` (DAA methods), 
`prop_signal` (signal percentages), `s2_vs` (variance parameters), and 
`ns` (sample sizes) to their desired values (e.g., `prop_signal=c(0.1, 0.2)`)
5. Run the script and use `process_output.R` to create figures.

In addition to our simulation study, we applied the proposed normalization methods 
to the PHACS and MLVS/MBS data. To replicate this analysis using synthetic versions of these datasets, run the file `analysis/real_data_analysis.R`, which will finish quickly. 



## Applying group-wise normalization

This section provides instructions for using group-wise normalization to analyze
real microbiome data. The first step is to generate fake versions
of PHACS and MLVS/MBS and to load the required functions. Please review the file
`utils/install_packages.R` to ensure you have installed the appropriate R packages.

```{r load_stuff, message=FALSE, warning=FALSE}
source("utils/generate_fake_data.R")
source("utils/general_sim_functions.R")
```

We will consider the task of performing DAA on MLVS/MBS. Examination of the
data reveals there are 372 taxa (the rows) measured on 520 subjects (the columns). 
The subjects have a mean sequencing depth (or "library size") of 61 thousand taxa
(range: 29 thousand to 110 thousand). Half the subjects have a value of 1 
for the covariate and half the subjects have 0. At present, the group-wise 
normalization framework can only be applied when the covariate of interest is binary. 
Note that, on real data, the library size will not equal the sample-wise
sum of counts since some taxa are typically omitted. Omitted taxa should still be
counted in the library size. 

```{r summarize_data}
load("data/MLVS_MBS.rda")
taxa <- mlvs_mbs$taxa
covariate <- mlvs_mbs$x
sequencing_depth <- colSums(taxa)
str(taxa)
table(covariate)
summary(sequencing_depth)

```

The main functions we will use are `groupnorm()` and `analysis_wrapper()`. 
The function `groupnorm()` calculates the corrected offset term by one of the two proposed
normalization methods, group-wise relative log expression (G-RLE) or 
fold-truncated sum scaling (FTSS). G-RLE is a simple normalization method that 
applies RLE, a traditional sample-wise normalization method, to the group-level 
log fold changes instead of their sample-level analogs. FTSS is a more complex 
method that selects reference taxa based on the proximity of their log fold changes 
to the mode log fold change. The method involves an additional hyper-parameter 
$p$, which denotes proportion of taxa that will serve as reference taxa. Our 
simulations demonstrated that FTSS is generally more effective at G-RLE at 
maintaining the nominal FDR level and offers slightly better power. Setting 
$p=0.4$ is a reasonable default.




```{r normalize}
# G-RLE
offset_grle <- groupnorm(Y = taxa, x = covariate, libsize = sequencing_depth,
                         method = "G-RLE")
summary(offset_grle)

# FTSS
offset_ftss <- 
  groupnorm(
    Y = taxa, x = covariate, 
    libsize = sequencing_depth,
    method = "FTSS", 
    prop_reference = 0.40 # Default value
  )
summary(offset_ftss)
```


The last step is to perform DAA using `analysis_wrapper()`. For simplicity, we
use the normalization method FTSS and the DAA methods 
MetagenomeSeq and DESeq2. 

`analysis_wrapper()` produces a data frame with the columns
`beta_1_hat` and `pv`, which contain the estimated log fold change and p-value
for each taxon, respectively. At a nominal FDR level of 5% based on the Benjamini and 
Hochberg procedure, MetagenomeSeq identified 23 taxa as associated with covariate
and DESeq2 identified 21.  


```{r mgs}
# MetagenomeSeq
out_mgs <-
  analysis_wrapper(
    Y = taxa, 
    x = covariate,
    offset = offset_ftss,
    method = "mgs"
  )
out_mgs |> 
  tibble() |> 
  mutate(
    taxon = paste0("taxon",1:n()),
    across(c(beta1_hat,pv), ~ signif(.x,3))
  ) |> 
  select(taxon, beta1_hat, pv) |> 
  filter(p.adjust(pv) < 0.05)

```      


```{r deseq}
# DESeq2
out_deseq <-
    analysis_wrapper(
    Y = taxa, 
    x = covariate,
    offset = offset_ftss,
    method = "DESeq2"
  )
out_deseq |>
  tibble() |> 
  mutate(
    taxon = paste0("taxon",1:n()),
    across(c(beta1_hat,pv), ~ signif(.x,3))
  ) |> 
  select(taxon, beta1_hat, pv) |> 
  filter(p.adjust(pv) < 0.05)


```  

To obtain additional results such as the estimated
standard error of the log fold changes, or for more complex analyses, use the 
`DESeq2::DESeq()` or `metagenomeSeq::fitFeatureModel()` functions that
`analysis_wrapper()` calls under the hood. Example code can be viewed by 
typing `View(analysis_wrapper)`. 



